### 2025-07-02 Text/Thesis/Chapters/Chapter2.qmd
update

In this work, we do not focus on parameter estimation using likelihood-based or moment-based methods. Instead, we adopt a testing framework to detect heterogeneity in SAR imagery. The analytical expressions of entropy measures derived from the $\Gamma_{\text{SAR}}$ and $\mathcal{G}_I^0$ models form the theoretical basis for constructing these statistical tests. In the following section, we introduce the entropy-based measures, specifically Shannon, Rényi, and Tsallis, on which our detection strategy is built.

@@

\textbf{Tsallis entropy},  was first introduced by Havrda and Charvat&nbsp;[-@Havrda1967]  in the context of information theory. Then, Tsallis&nbsp;[-@Tsallis1988] exploited its non-extensive features and placed it in a physical setting, known as generalized Tsallis entropy of order $\lambda$,  offers an alternative generalization motivated by non-extensive systems, those exhibiting long-range dependencies, memory effects, or multifractal behavior. Unlike Rényi, Tsallis entropy does not obey the additivity property under independent systems, which makes it particularly suitable for modeling complex spatial structures and textured regions in SAR images.

@@

<!-- When $\alpha\!\to\!-\infty$ (fully developed speckle), this excess tends to zero, -->
<!--  so the heterogeneous model collapses to the homogeneous one in Equation \eqref{eq-HGammaSAR}.   -->
<!-- and detecting heterogeneity is equivalent to deciding whether $\Delta_\alpha$ is significantly different from zero. -->
<!--  Note that the Rényi entropy of $\mathcal{G}^0_I$ can be expressed in terms of that of $\Gamma_{\mathrm{SAR}}$, with additional terms that depend on the roughness parameter $\alpha$. As $\alpha \to -\infty$, these terms tend to zero, reducing&nbsp;\eqref{eq-HGI0} to&nbsp;\eqref{eq-HGammaSAR}.  -->
<!-- This work's core idea is to check if this excess entropy is significantly different from zero. -->
<!-- Fig.&nbsp;\ref{fig-convergence-renyi} illustrates this behaviour: as $\alpha$ decreases the colored curves of $R_\lambda(\mathcal{G}^0_I)$ converge towards the solid black curve of $R_\lambda(\Gamma_{\mathrm{SAR}})$, confirming the limiting behavior. -->

>>>>

In this work, we do not focus on parameter estimation using likelihood-based or moment-based methods. Instead, we adopt a hypothesis testing framework to detect heterogeneity in SAR imagery. The analytical expressions of entropy measures derived from the $\Gamma_{\text{SAR}}$ and $\mathcal{G}_I^0$ models form the theoretical basis for constructing these statistical tests. In the following section, we introduce the entropy-based measures, specifically Shannon, Rényi, and Tsallis, on which our detection strategy is built.

@@

\textbf{Tsallis entropy} was first introduced by Havrda and Charvát&nbsp;[-@Havrda1967] in the context of information theory. Later, Tsallis&nbsp;[-@Tsallis1988] extended it by emphasizing its non-extensive features and placing it in a physical context. The generalized Tsallis entropy of order $\lambda$ offers an alternative formulation motivated by non-extensive systems, such as those exhibiting long-range dependencies, memory effects, or multifractal behavior. Unlike Rényi entropy, it does not obey the additivity property under independence, making it particularly suitable for modeling complex spatial structures and textured regions in SAR images.

@@

### 2025-07-02 Text/Thesis/Chapters/Chapter3.qmd
update

Figure \ref{fig:Plot_bias_mse-sh} presents the bias and MSE for each of the non-parametric estimators of Shannon entropy. The results are summarized in @tbl-out.

@@

\renewcommand{\arraystretch}{1.8} 
```{r Table_estimators_shannon2, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
#| label: tbl-out
#| tbl-cap: 'Bias and MSE of Shannon entropy estimators under $\Gamma_{\text{SAR}}(5,1)$.'

@@

The bias is then estimated using Equation \eqref{eq:bias_estimate}, which calculates the difference between the average of the bootstrap estimates and the original estimate:
\begin{equation}
\widehat{\operatorname{Bias}} = \frac{1}{B} \sum_{b=1}^{B} \widehat{\theta}(\bm{Z}^{(b)}) - \widehat{\theta}(\bm{Z}).
\label{eq:bias_estimate}
\end{equation}
A bias-corrected estimator is then defined by subtracting the estimated bias from the original estimator, as expressed in Equation \eqref{eq:bias_corrected}:
\begin{equation}
\widetilde{\theta} = \widehat{\theta} - \widehat{\operatorname{Bias}}.
\label{eq:bias_corrected}
\end{equation}
Substituting the estimate from Equation \eqref{eq:bias_estimate} into Equation \eqref{eq:bias_corrected} yields the explicit formula for the corrected estimator:
\begin{equation}
\widetilde{\theta} = 2\widehat{\theta}(\bm{Z}) - \frac{1}{B} \sum_{b=1}^{B} \widehat{\theta}(\bm{Z}^{(b)}).
\label{eq:theta_explicit}
\end{equation}
<!-- This expression follows from the classical bias correction formula: -->
<!-- \begin{equation} -->
<!-- \text{Bias-corrected estimate} = \text{Original estimate} - \text{Bias estimate}. -->
<!-- \label{eq:bias_correction_principle} -->
<!-- \end{equation} -->
Thus, $\widetilde{\theta}$ is an approximately unbiased estimator of $\theta$. The effectiveness of the correction depends on the number of bootstrap replicates $B$ and the characteristics of the underlying distribution, but it systematically reduces bias, particularly when working with small sample sizes.

@@

 @tbl-out-time shows the processing time of each estimator across different settings of $L$  and $n$.
<!-- <!-- # ```{r Simulated_comparative_time6, eval=FALSE, echo=FALSE} --> 
```{r Simulated_comparative_time6, echo=FALSE, message=FALSE}
set.seed(1234567890, kind = "Mersenne-Twister")
R <- 600
mu <- 1
B <- 40

@@

```
\renewcommand{\arraystretch}{1.1}
```{r Table_statistic_time_6,  echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
#| label: tbl-out-time
#| tbl-cap: 'Processing time for bootstrap-improved estimators $\widetilde{H}_{\text{C}}$ and $\widetilde{H}_{\text{AO}}$'

@@

<!-- ```{r Simulated_data_bias_shannon1, echo=FALSE, message=FALSE} -->
<!-- set.seed(1234567890) -->

<!-- file_name <- "./Data/results_shannon_v1.Rdata" -->

<!-- if (file.exists(file_name)) { -->
<!--   load(file_name) -->
<!--   message("Loaded existing results from results_shannon_v1.Rdata") -->
<!-- } else { -->
<!--   sample_sizes <- c(9, 25, 49, 81, 121) -->
<!--   R <- 500 -->
<!--   B <- 100 -->
<!--   mu <- 1 -->
<!--   L <- 16 -->

<!--   estimators <- list( -->
<!--     "Shannon Estimator" = al_omari_1_estimator, -->
<!--     "Shannon Estimator Bootstrap" = bootstrap_al_omari_1_estimator -->
<!--   ) -->

<!--   results <- calculate_bias_mse_shannon(sample_sizes, R, B, mu, L, estimators) -->

<!--   save(results, file = file_name) -->
<!--   message("Simulaciones completadas y resultados guardados.") -->
<!-- } -->

<!-- ``` -->
<!-- ```{r fig-bias_mse_shannonm, echo=FALSE, message=FALSE, warning=FALSE, out.width="80%", fig.cap="Bias and MSE of the Shannon–Al-Omari entropy estimators for $\\Gamma_{\\text{SAR}}$.", fig.width=6, fig.height=5} -->
<!-- load("./Data/results_shannon_v1.Rdata") -->

<!-- estimators_to_plot <- c("Shannon Estimator", "Shannon Estimator Bootstrap") -->
<!-- latex_estimator_names <- c( -->
<!--   "Shannon Estimator" = expression("$\\widehat{italic(H)}$"), -->
<!--   "Shannon Estimator Bootstrap" = expression("$\\widetilde{italic(H)}$") -->
<!-- ) -->
<!-- selected_estimators_latex <- latex_estimator_names[estimators_to_plot] -->

<!-- combined_plot_shannon <- generate_plot_shannon(results, selected_estimators_latex, ncol = 1, nrow = 1) -->

<!-- print(combined_plot_shannon) -->

<!-- ``` -->

<!-- ```{=latex} -->
<!-- \begin{figure}[hbt] -->
<!--   \centering -->
<!--   \includegraphics[width=0.40\textwidth]{Figures-R1/fig-bias_mse-1.pdf} -->
<!--   \vspace{0.1em} -->
<!--   \caption{Bias and MSE of the Rényi entropy estimators for $\Gamma_{\text{SAR}}$.} -->
<!--   \label{fig-bias_mse} -->
<!-- \end{figure} -->
<!-- ``` -->
<!-- We use the $\widetilde{H}_{\lambda}$ estimator for subsequent simulations. -->

@@

  # p <- ggplot(combined_data, aes(x = Test_Statistics, col = LegendOrder, linetype = LegendOrder, group = CurveOrder)) +
  #   geom_line(stat = "density", linewidth = 1.5) +
  #   scale_color_viridis(discrete = TRUE, option = "C", direction = -1, begin = 0.2, end = 0.7, name = "Sample Size") +
  #   scale_linetype_manual(values = rep("solid", length(sample.size)), name = "Sample Size") +
  #   scale_x_continuous(limits = x_limits, breaks = x_breaks, minor_breaks = NULL) +
  #   scale_y_continuous(limits = y_limits, minor_breaks = NULL) +
  #   labs(
  #       x = expression("Test Statistic" ~ S[widetilde(italic(T))](italic(bold(Z))* ";" **phantom(" ")* italic(L))), 
  #       y = "Density"
  #   ) +
  #   ggtitle(bquote(italic(L) == .(L))) +
  #   theme(plot.title = element_text(hjust = 0.5, size = 16, margin = margin(b = 2)),
  #         axis.text = element_text(size = 16),     
  #         axis.title = element_text(size = 16),    
  #         legend.text = element_text(size = 16),   
  #         legend.title = element_text(size = 16))

@@

  # p <- ggplot(combined_data, aes(x = Test_Statistics, col = LegendOrder, linetype = LegendOrder, group = CurveOrder)) +
  #   geom_line(stat = "density", linewidth = 1.5) +
  #   scale_color_viridis(discrete = TRUE, option = "C", direction = -1, begin = 0.2, end = 0.7, name = "Sample Size") +
  #   scale_linetype_manual(values = rep("solid", length(sample.size)), name = "Sample Size") +
  #   scale_x_continuous(limits = x_limits, breaks = x_breaks, minor_breaks = NULL) +
  #   #scale_x_continuous(limits = x_limits, breaks = x_breaks) + 
  #   scale_y_continuous(limits = y_limits, minor_breaks = NULL)+
  #   #scale_y_continuous(limits = y_limits) +  
  #   labs(
  #       x = expression("Test Statistic" ~ S[widetilde(italic(T))[lambda]](italic(bold(Z))* ";" **phantom(" ")* italic(L))), 
  #       y = "Density"
  #   ) +
  #   ggtitle(bquote(italic(L) == .(L))) +
  #   theme(plot.title = element_text(hjust = 0.5, size = 16, margin = margin(b = 2)),
  #         axis.text = element_text(size = 16),     
  #       axis.title = element_text(size = 16),    
  #       legend.text = element_text(size = 16),   
  #       legend.title = element_text(size = 16)   
  #         )

@@

The size of a statistical test, also known as the Type&nbsp;I error rate, refers to the probability of incorrectly rejecting the null hypothesis $\mathcal{H}_0$ when it is in fact true. In hypothesis testing, practitioners typically specify a nominal significance level, commonly set at \SI{1}{\percent}, \SI{5}{\percent}, and \SI{10}{\percent}.  These values define the acceptable probability of committing a Type&nbsp;I error. A well-calibrated test should exhibit empirical Type~I error rates that closely match the nominal levels across different sample sizes and conditions.

>>>>

Figure \ref{fig:Plot_bias_mse-sh} presents the bias and MSE for each of the non-parametric estimators of Shannon entropy. The results are summarized in @tbl-out-shannon.

@@

\renewcommand{\arraystretch}{1.8} 
```{r}
#| label: tbl-out-shannon
#| echo: false
#| message: false
#| warning: false
#| results: 'asis'
#| tbl-cap: 'Bias and MSE of Shannon entropy estimators under $\Gamma_{\text{SAR}}(5,1)$.'

@@

The bias is then estimated using Equation \eqref{eq:bias_estimate}, which calculates the difference between the average of the bootstrap estimates and the original estimate:
\begin{equation}
\widehat{\operatorname{Bias}} = \frac{1}{B} \sum_{b=1}^{B} \widehat{\theta}(\bm{Z}^{(b)}) - \widehat{\theta}(\bm{Z}).
\label{eq:bias_estimate}
\end{equation}
A bias-corrected estimator is then defined by subtracting the estimated bias from the original estimator, as expressed in Equation \eqref{eq:bias_corrected}:
\begin{equation}
\widetilde{\theta} = \widehat{\theta} - \widehat{\operatorname{Bias}}.
\label{eq:bias_corrected}
\end{equation}
Substituting the estimate from Equation \eqref{eq:bias_estimate} into Equation \eqref{eq:bias_corrected} yields the explicit formula for the corrected estimator:
\begin{equation}
\widetilde{\theta} = 2\widehat{\theta}(\bm{Z}) - \frac{1}{B} \sum_{b=1}^{B} \widehat{\theta}(\bm{Z}^{(b)}).
\label{eq:theta_explicit}
\end{equation}
Thus, $\widetilde{\theta}$ is an approximately unbiased estimator of $\theta$. The effectiveness of the correction depends on the number of bootstrap replicates $B$ and the characteristics of the underlying distribution, but it systematically reduces bias, particularly when working with small sample sizes.

@@

 @tbl-out-time shows the processing time of each estimator across different settings of $L$  and $n$.

```{r Simulated_comparative_time6, echo=FALSE, message=FALSE}
set.seed(1234567890, kind = "Mersenne-Twister")
R <- 600
mu <- 1
B <- 40

@@

\renewcommand{\arraystretch}{1.1}
```{r}
#| label: tbl-out-time
#| echo: false
#| message: false
#| warning: false
#| results: 'asis'
#| tbl-cap: 'Processing time for bootstrap-improved estimators $\widetilde{H}_{\text{C}}$ and $\widetilde{H}_{\text{AO}}$'

@@

@@

@@

@@

The size of a statistical test, also known as the Type&nbsp;I error rate, refers to the probability of incorrectly rejecting the null hypothesis $\mathcal{H}_0$ when it is in fact true. In hypothesis testing, practitioners typically specify a nominal significance level, commonly set at \SI{1}{\percent}, \SI{5}{\percent}, and \SI{10}{\percent}.  These values define the acceptable probability of committing a Type&nbsp;I error. A well-calibrated test should exhibit empirical Type&nbsp;I error rates that closely match the nominal levels across different sample sizes and conditions.

### 2025-07-02 Text/Thesis/Chapters/Chapter4.qmd
update

Eight polygonal regions of interest (ROIs) were manually selected for each of the three SAR scenes chosen for quantitative evaluation, London, Munich, and Dublin (Figures \ref{fig:london-sar-roi1}–\ref{fig:dublin-sar-roi1}). These ROIs correspond to heterogeneous areas (red, class 1) and homogeneous areas (blue, class 0). The polygons were rasterized to match the spatial resolution of each SAR image, forming a partial ground truth. We thresholded the $p$-value maps at 0.05, labeling each pixel as heterogeneous (1) or homogeneous (0), and compared these decision maps with the ground truth to compute the F1-score ($F_1$), Kappa coefficient ($\kappa$), and overall accuracy (OA). Table \ref{tab:table_metrics1} summarizes the results.

```{=latex}
\begin{figure}[H]
    \centering
    % London
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figures/london_roi.png}
        \caption{London SAR image with selected ROIs.}
        \label{fig:london-sar-roi1}
    \end{subfigure}
   \hspace{4em} 
    % Munich
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figures/munich_roi.png}
        \caption{Munich SAR image with selected ROIs.}
        \label{fig:munich-sar-roi1}
    \end{subfigure}
    \vspace{1.8em}
    % Dublin
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figures/dublin_roi3.png}
        \caption{Dublin SAR image with selected ROIs.}
        \label{fig:dublin-sar-roi1}
    \end{subfigure}
    \caption{Selected ROIs for the quantitative evaluation. Red polygons indicate heterogeneous areas (class 1), and blue polygons indicate homogeneous areas (class 0).}
    \label{fig:roi-scenes}
\end{figure}

When comparing the three entropy-based tests, distinct patterns emerge depending on the scene characteristics. In London, the Rényi-based test produced the best performance across all metrics ($F_1$: 0.695; $\kappa$: 0.626; OA: 0.877), outperforming both Shannon and Tsallis. This suggests that Rényi entropy is particularly robust under strong speckle conditions, as is typical in single-look SAR imagery.

@@

>>>>

Eight polygonal regions of interest (ROIs) were manually selected for each of the three SAR scenes chosen for quantitative evaluation, London, Munich, and Dublin (Figures \ref{fig:london-sar-roi1}–\ref{fig:dublin-sar-roi1}). These ROIs correspond to heterogeneous areas (red, class 1) and homogeneous areas (blue, class 0). The polygons were rasterized to match the spatial resolution of each SAR image, forming a partial ground truth.

```{=latex}
\begin{figure}[H]
    \centering
    % London
    \begin{subfigure}{0.30\textwidth}
        \includegraphics[width=\linewidth]{Figures/london_roi.png}
        \caption{London SAR image with selected ROIs.}
        \label{fig:london-sar-roi1}
    \end{subfigure}
   \hspace{4em} 
    % Munich
    \begin{subfigure}{0.30\textwidth}
        \includegraphics[width=\linewidth]{Figures/munich_roi.png}
        \caption{Munich SAR image with selected ROIs.}
        \label{fig:munich-sar-roi1}
    \end{subfigure}
    \vspace{0.5em}
    % Dublin
    \begin{subfigure}{0.30\textwidth}
        \includegraphics[width=\linewidth]{Figures/dublin_roi3.png}
        \caption{Dublin SAR image with selected ROIs.}
        \label{fig:dublin-sar-roi1}
    \end{subfigure}
    \caption{Selected ROIs for the quantitative evaluation. Red polygons indicate heterogeneous areas (class 1), and blue polygons indicate homogeneous areas (class 0).}
    \label{fig:roi-scenes}
\end{figure}

```
We thresholded the $p$-value maps at 0.05, labeling each pixel as heterogeneous (1) or homogeneous (0), and compared these decision maps with the ground truth to compute the F1-score ($F_1$), Kappa coefficient ($\kappa$), and overall accuracy (OA). Table \ref{tab:table_metrics1} summarizes the results.

@@

``` 
When comparing the three entropy-based tests, distinct patterns emerge depending on the scene characteristics. In London, the Rényi-based test produced the best performance across all metrics ($F_1$: 0.695; $\kappa$: 0.626; OA: 0.877), outperforming both Shannon and Tsallis. This suggests that Rényi entropy is particularly robust under strong speckle conditions, as is typical in single-look SAR imagery.

### 2025-07-02 Text/Thesis/Frontmatter/abstract.tex
update

This work  presents a statistical approach to detect heterogeneity 
in synthetic aperture radar (SAR) intensity data using entropy-based methods.
In homogeneous regions, where fully developed speckle is present, SAR returns 
are well modeled by the Gamma distribution. 
In contrast, heterogeneous areas exhibit more complex scattering behaviors, 
typically modeled by the $\mathcal{G}_I^0$ distribution. Distinguishing between 
these regimes is fundamental for accurate interpretation in remote sensing applications, 
yet classical parametric tests often fail to address this task effectively due to analytical 
and numerical limitations.
To overcome these challenges, we propose three test statistics for detecting heterogeneity 
in SAR imagery based on Shannon, Rényi, and Tsallis entropy measures. These tests rely on 
nonparametric entropy estimators constructed from sample spacings, avoiding explicit assumptions 
about the underlying distribution. To enhance their accuracy, particularly in small samples, 
we incorporate a bootstrap bias-correction procedure that improves the estimators’ stability, 
reduces bias, and lowers mean squared error.
The proposed tests are evaluated using Monte Carlo simulations, assessing size and power 
under varying sample sizes and numbers of looks.
Results demonstrate that the Rényi and Tsallis-based tests outperform the Shannon-based version by 
detecting more subtle texture variations while maintaining greater reliability in identifying 
truly homogeneous regions.
Finally, the methodology is applied to both simulated and real SAR data. We generate $p$-value maps using a 
sliding window analysis, allowing for visual and quantitative assessment of spatial heterogeneity. 
The Rényi-based test consistently identifies fine-scale roughness patterns, 
while the Tsallis-based test shows strength in reliably detecting homogeneous areas. 
Together, these entropy-based tools provide a robust, interpretable, and unsupervised 
framework for heterogeneity detection in SAR data.

\vspace{1em}
\par
\noindent \textbf{Keywords:} SAR; Heterogeneity; Entropy; Nonparametric Estimation; Bootstrap; Statistical Test

>>>>

This work presents a statistical approach to detect heterogeneity 
in synthetic aperture radar (SAR) intensity data using entropy-based methods.
In SAR data analysis, accurate terrain interpretation fundamentally depends 
on distinguishing two key regimes: homogeneous regions, characterized by fully 
developed speckle and modeled by the Gamma distribution, and heterogeneous areas, 
which exhibit complex scattering behaviors typically captured by the $\mathcal{G}_I^0$ distribution.
While this discrimination is essential for remote sensing applications, classical parametric tests often fail to address it 
effectively due to analytical limitations and numerical instabilities.
To overcome these challenges, we propose three test statistics for detecting heterogeneity 
in SAR imagery based on Shannon, Rényi, and Tsallis entropy measures. These tests rely on 
nonparametric entropy estimators constructed from sample spacings, avoiding explicit assumptions 
about the underlying distribution. To enhance their accuracy, particularly in small samples, 
we incorporate a bootstrap bias-correction procedure that improves the stability of the estimators, 
reduces bias, and lowers mean squared error.
The proposed tests are evaluated using Monte Carlo simulations, assessing size and power 
under varying sample sizes and numbers of looks.
Results demonstrate that the Rényi and Tsallis-based tests outperform the Shannon-based version by 
detecting more subtle texture variations while maintaining greater reliability in identifying 
truly homogeneous regions.
Finally, the methodology is applied to both simulated and real SAR data. We generate $p$-value maps using a 
sliding window analysis, allowing for visual and quantitative assessment of spatial heterogeneity. 
The Rényi-based test consistently identifies fine-scale roughness patterns, while the Tsallis-based 
test proves more effective in reliably detecting homogeneous areas. Together, these entropy-based 
tools provide a robust, interpretable, and fully unsupervised framework for heterogeneity detection 
in SAR data.

\vspace{1em}
\par
\noindent \textbf{Keywords:}  Entropy; SAR; Heterogeneity; Gamma distribution; Hypothesis testing; Bootstrap

### 2025-07-02 Text/Thesis/Frontmatter/resumo.tex
update

Este trabalho apresenta uma abordagem estatística para a detecção de heterogeneidade 
em dados de intensidade de radar de abertura sintética (SAR) utilizando medidas 
baseadas em entropia. Em regiões homogêneas, onde há speckle totalmente desenvolvido, 
os retornos SAR são bem modelados pela distribuição Gama. Em contraste, áreas heterogêneas 
apresentam comportamentos de espalhamento mais complexos, normalmente descritos pela 
distribuição $\mathcal{G}_I^0$. Distinguir entre esses dois regimes é fundamental 
para uma interpretação precisa em aplicações de sensoriamento remoto. No entanto, 
testes paramétricos clássicos geralmente não são adequados para essa tarefa devido a 
limitações analíticas e numéricas.
Para superar esses desafios, propomos três estatísticas de teste para detectar heterogeneidade 
em imagens SAR com base nas entropias de Shannon, Rényi e Tsallis. Esses testes utilizam 
estimadores não paramétricos de entropia construídos a partir de espaçamentos amostrais, 
evitando suposições explícitas sobre a distribuição subjacente. Para aumentar a precisão dos testes, 
especialmente em amostras pequenas, incorporamos um procedimento de correção de viés via bootstrap, 
que melhora a estabilidade dos estimadores, reduz o viés e o erro quadrático médio.
As estatísticas propostas são avaliadas por meio de simulações de Monte Carlo, 
onde analisamos seu tamanho e poder sob diferentes condições de speckle e textura. 
Os resultados mostram que os testes baseados em Rényi e Tsallis superam a versão baseada 
em Shannon, detectando variações de textura mais sutis e mantendo maior confiabilidade na 
identificação de regiões verdadeiramente homogêneas.
Por fim, a metodologia é aplicada a dados SAR simulados e reais. 
A análise é realizada com janelas deslizantes, gerando mapas de valores de $p$ que permitem a a
valiação visual e quantitativa da heterogeneidade espacial. O teste baseado na entropia de 
Rényi mostra desempenho superior na identificação de padrões de rugosidade em pequena escala, 
enquanto o teste baseado em Tsallis é eficaz na detecção de regiões homogêneas. Em conjunto, 
essas ferramentas baseadas em entropia oferecem uma estrutura robusta, interpretável e não 
supervisionada para a detecção de heterogeneidade em dados SAR.

\vspace{1em}
\par
\noindent \textbf{Palavras-chave:} SAR; Heterogeneidade; Entropia; Estimativa não paramétrica; Bootstrap; Teste estatístico

>>>>

Este trabalho apresenta uma abordagem estatística para detectar heterogeneidade em dados de 
intensidade de radar de abertura sintética (SAR) utilizando métodos baseados em entropia. 
Na análise de dados SAR, uma interpretação precisa do terreno depende fundamentalmente da 
distinção entre dois regimes principais: regiões homogêneas, caracterizadas por speckle 
totalmente desenvolvido e modeladas pela distribuição Gamma, e áreas heterogêneas, que 
exibem comportamentos de espalhamento complexos, geralmente representados pela distribuição 
$\mathcal{G}_I^0$.
Embora essa discriminação seja essencial para aplicações de sensoriamento remoto, 
testes paramétricos clássicos muitas vezes não a abordam de forma eficaz devido a 
limitações analíticas e instabilidades numéricas.
Para superar esses desafios, propomos três estatísticas de teste para detectar heterogeneidade 
em imagens SAR com base nas entropias de Shannon, Rényi e Tsallis. Esses testes utilizam 
estimadores não paramétricos de entropia construídos a partir de espaçamentos amostrais, 
evitando suposições explícitas sobre a distribuição subjacente. Para aumentar a precisão dos testes, 
especialmente em amostras pequenas, incorporamos um procedimento de correção de viés via bootstrap, 
que melhora a estabilidade dos estimadores, reduz o viés e o erro quadrático médio.
As estatísticas propostas são avaliadas por meio de simulações de Monte Carlo, 
onde analisamos seu tamanho e poder sob diferentes condições de speckle e textura. 
Os resultados mostram que os testes baseados em Rényi e Tsallis superam a versão baseada 
em Shannon, detectando variações de textura mais sutis e mantendo maior confiabilidade na 
identificação de regiões verdadeiramente homogêneas.
Por fim, a metodologia é aplicada a dados SAR simulados e reais. 
A análise é realizada com janelas deslizantes, gerando mapas de valores de $p$ que permitem a a
valiação visual e quantitativa da heterogeneidade espacial. O teste baseado na entropia de 
Rényi mostra desempenho superior na identificação de padrões de rugosidade em pequena escala, 
enquanto o teste baseado em Tsallis é eficaz na detecção de regiões homogêneas. Em conjunto, 
essas ferramentas baseadas em entropia oferecem uma estrutura robusta, interpretável e não 
supervisionada para a detecção de heterogeneidade em dados SAR.

\vspace{1em}
\par
\noindent \textbf{Palavras-chave:} Entropia; SAR; Heterogeneidade; Distribuição Gama; Teste de hipótese; Bootstrap
